---
title: "DE_limma_FW_v_BW"
author: "Lisa Johnson / Joe McGirr"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    collapsed: no
    df_print: paged
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_depth: 5
    toc_float: yes
  html_notebook:
    toc: yes
    toc_depth: 5
editor_options: 
  chunk_output_type: console
---

# Introduction

A mixed model with random effects was chosen for this multifactor experiment, and analyzed using the `limma` package in R. This package implements a linear modeling approach and empirical Bayes statistics. The `limma` package with the `voom` method estimates the mean-variance relationship of the log-counts, generates a precision weight for each observation and enters these into the limma empirical Bayes analysis pipeline.

In this model, `clade` (levels = Clade1, Clade2, Clade3), `physiology` (levels = Marine, Freshwater), and experimental `condition` (levels = 15ppt, 0.2ppt) are fixed effects while `species` (levels = 14) is considered a random effect.

https://www.bioconductor.org/packages/devel/workflows/vignettes/RNAseq123/inst/doc/limmaWorkflow.html
https://physiology.med.cornell.edu/faculty/skrabanek/lab/angsd/lecture_notes/Background_testingForDGE.pdf

```{r LoadPackages, results='hide', include=FALSE, warning=FALSE}

# Install function for packages    
packages<-function(x){
  x<-as.character(match.call()[[2]])
  if (!require(x,character.only=TRUE)){
    install.packages(pkgs=x,repos="http://cran.r-project.org")
    require(x,character.only=TRUE)
  }
}


packages(MASS)
packages(ggplot2)
packages(gtools)
packages(pheatmap)
packages(cowplot)
packages(RColorBrewer)
packages(dplyr)
packages(tidyr)
packages(knitr)
packages(ggrepel)
packages(DESeq2)
packages(limma)
packages(edgeR)
packages(gplots)
packages(lattice)
packages(vsn)
packages(biomaRt)
packages(kableExtra)
packages(pheatmap)
packages(SummarizedExperiment)
packages(emmeans)
packages(data.table)
packages(ggpubr)

# color-blind friendly 
# Wong, B. Points of view: Color blindness. Nat Methods (2011).
bla <- "#000000"
blu <- "#0072b2"
grb <- "#56b4e9"
lir <- "#cc79a7"
gre <- "#009e73"
red <- "#d55e00"
org <- "#e69f00"
yel <- "#f0e442"
gry<-  '#BBBBBB'
######

```

The raw counts file, generated with `NumReads` from the salmon (version 0.12.0) quantification tool and summarized with the tximport Bioconductor package (version 1.10.1) in R, can be downloaded from an [osf repository](https://osf.io/m4xeg/) with this [link](https://osf.io/7vp38/download), then imported into the R framework.

```{r loadfiles, results='hide', include=FALSE, warning=FALSE}
# This is the counts with Experimental Design Info in the last 5 rows

counts_design <- read.csv("C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/expression_tables/Ensembl_species_counts_designfactors.csv",stringsAsFactors = FALSE)

```

Samples from species with low numbers of replicates were dropped from the raw counts table (*F. zebrinus*, *F. nottii*, *F. sciadicus*). The raw counts table has the following dimensions (genes x samples). 

```{r dropsamples,results='show', warning=FALSE}

#dim(counts_design)
#[1] 31595   130

# -----------------------
# Determine which columns to drop
# based on sample size
# Drop columns with no data
# -----------------------

length(counts_design)
samps <- as.data.frame(colnames(counts_design)[3:length(counts_design)])
samps <- samps %>% separate(`colnames(counts_design)[3:length(counts_design)]`, , into = c("species", "genus", "treatment","replicate"), sep = "_")
samps$group <- paste(samps$species,samps$genus,samps$treatment, sep = ":")
table(samps$group)
length(unique(paste(samps$species,samps$genus)))

samps_t <- samps %>% filter(treatment == "transfer")
samps_b <- samps %>% filter(treatment == "BW")
samps_f <- samps %>% filter(treatment == "FW")

length(unique(paste(samps_t$species,samps_t$genus, samps_t$treatment)))
length(unique(paste(samps_b$species,samps_b$genus, samps_b$treatment)))
length(unique(paste(samps_f$species,samps_f$genus, samps_f$treatment)))
setdiff(unique(paste(samps_f$species,samps_f$genus)),
        unique(paste(samps_t$species,samps_t$genus)))

design <- counts_design[counts_design$Ensembl == 'Empty',]
#design$type <- c("species","native_salinity","clade","group","condition")

# dropping all zebrinus, nottii, sciadicus, due to sample size
drops <- c("X","Ensembl",
           "F_zebrinus_BW_1.quant","F_zebrinus_BW_2.quant",
           "F_zebrinus_FW_1.quant","F_zebrinus_FW_2.quant",
           "F_nottii_FW_1.quant","F_nottii_FW_2.quant",
           "F_sciadicus_BW_1.quant","F_sciadicus_FW_1.quant",
           "F_sciadicus_FW_2.quant","F_sciadicus_transfer_1.quant")

bw_drops <- grep('_BW_', colnames(counts_design), value=TRUE)

counts<-counts_design[!counts_design$Ensembl == 'Empty',]
rownames(counts)<-counts$Ensembl
design <- design[ , !(names(design) %in% drops)]
counts <- counts[ , !(names(counts) %in% drops)]
design <- design[ , !(names(design) %in% bw_drops)]
counts <- counts[ , !(names(counts) %in% bw_drops)]
dim(design)
dim(counts)
gene.names<-rownames(counts)
design[] <- lapply( design, factor)
```

# Sample Design Matrix

A matrix was generated using the following model with fixed effects: 

```
~physiology*condition*clade
```
The random effect of `species` will be taken into account later.

```{r designinfo,results='show', warning=FALSE}

# --------------------
# design categories
# --------------------

species<-as.character(unlist(design[1,]))
physiology<-as.character(unlist(design[2,]))
clade<-as.character(unlist(design[3,]))
condition<-as.character(unlist(design[5,]))
condition_physiology<-as.vector(paste(condition,physiology,sep="."))
condition_physiology_clade <- as.vector(paste(condition_physiology,clade,sep="."))
condition_physiology_clade <- as.vector(paste("group",condition_physiology_clade,sep=""))
condition_species <- as.vector(paste(condition,species,sep="."))
cols<-colnames(counts)
ExpDesign <- data.frame(row.names=cols,
                        condition=condition,
                        physiology = physiology,
                        clade = clade,
                        species = species,
                        sample=cols)
ExpDesign
# used for pairwise contrasts
#form<-as.formula("~0 + physiology*condition*clade")
form<-as.formula("~physiology*condition*clade")
design = model.matrix(form, ExpDesign)
#group <- interaction(physiology, condition, clade)
#mm <- model.matrix(~0 + group)
colnames(design)
# check rank of matrix
Matrix::rankMatrix( design )
dim(design)
clade <- ExpDesign$clade
physiology <- ExpDesign$physiology

#write.table(ExpDesign,"C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/DE/counts_threshold_new/ExpDesign.txt",sep="\t",quote=FALSE, row.names = FALSE)

```

# Filtering and Normalization

Genes with low expression across samples were dropped from the analysis using a conservative approach. The function `filterByExpr` was used on the raw counts matrix. For each `condition_physiology` group (regardless of species), each sample must have a minium count of 10, and a group minimum total count of 100. This reduced the counts table to the following dimensions (genes x samples):

```{r filt, results="show",warning=FALSE}

counts<-as.matrix(as.data.frame(sapply(counts, as.numeric)))
rownames(counts)<-gene.names
class(counts)
#test<-counts %>% drop_na()
#test<-as.matrix(test)
lcom_unfilt<-log2(counts+1)
plot(colSums(t(lcom_unfilt)))

#lisa's minimun count filter
#keep<-filterByExpr(counts,design = design,group=condition_physiology,min.count = 10, min.total.count = 100)
#joe's minimum count filter
keep<-filterByExpr(counts,design = design,group=condition_physiology_clade,min.count = 10, min.total.count = 30)

counts.filt <- counts[keep,]
dim(counts.filt)
#write.table(counts.filt,"C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/DE/counts_threshold_new/DE_limma_FW_v_transfer/exp.tsv",sep="\t",quote=FALSE)


```

# biomaRt annotation

```{r filt, results="show",warning=FALSE}
# ============================================
# biomart annotation
# https://uswest.ensembl.org/Fundulus_heteroclitus/Info/Index
# ============================================

ensembl=useMart("ENSEMBL_MART_ENSEMBL")
ensembl = useDataset("fheteroclitus_gene_ensembl",mart=ensembl)
# test
# id <- c("ENSGMOP00000000001.1","ENSGMOP00000000002.1","ENSGMOP00000000003.1")
ensembl_proteinID <- rownames(counts)
ensembl_proteinID <- unlist(strsplit(ensembl_proteinID,split="[.]1"))

length(ensembl_proteinID)
# can take >5 min
# do only once
ann<-getBM(attributes=c('ensembl_peptide_id','ensembl_transcript_id','ensembl_gene_id','gene_biotype','external_gene_name','description'), filters = 'ensembl_peptide_id', values = ensembl_proteinID, mart=ensembl)
head(ann)
dim(ann)
length(unique(ann$ensembl_peptide_id))
ann <- ann[!duplicated(ann[,c(1)]),]
dim(ann)
#write.table(ann,"C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/DE/counts_threshold_new/DE_limma_FW_v_transfer/annotations.tsv",sep="\t",quote=FALSE)

```


# DE analyses

```{r filt, results="show",warning=FALSE}

# log counts before DE
boxplot(log(counts.filt+1), las = 2, main = "")

genes = DGEList(count = counts.filt, group = condition_physiology_clade)
genes = calcNormFactors(genes)

# write counts per million normalized counts
tmp <- as.data.frame(cpm(genes))
tmp$Ensembl <- rownames(tmp)
tmp <- dplyr::select(tmp, Ensembl, everything())
#write.csv(tmp, file = file.path("C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/DE/counts_threshold_new/DE_limma_FW_v_transfer/normalized_counts.csv"), quote = F, row.names = F)

norm_counts <- read.csv("C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/DE/counts_threshold_new/DE_limma_FW_v_transfer/normalized_counts.csv",stringsAsFactors = FALSE, header = TRUE,row.names = "Ensembl")

dim(norm_counts)
# fix rownames to chop .1 off
rownames(norm_counts) <- unlist(strsplit(rownames(norm_counts),split="[.]1"))

vobj = voom( genes, design, plot=TRUE)
lcpm <- cpm(genes$counts, log = TRUE)

# log counts after DE

boxplot(lcpm, las = 2, main = "")
plot(colSums(t(lcpm)))

vwts <- voomWithQualityWeights(genes, design=design, normalization="quantile", plot=TRUE)
corfit <- duplicateCorrelation(vobj,design,block=ExpDesign$species)
corfit$consensus
#[1] 0.7213008

# species random effect
fitRan <- lmFit(vobj,design,block=ExpDesign$species,correlation=corfit$consensus)


```

# PCA

This is the PCA with all counts, not filtered. The dimensions of the counts table are listed below. Row=genes, Columns=samples  
normal full counts

```{r filt, results="show",warning=FALSE}
# -----------------
# PCA, all counts
# -----------------

# This is the PCA with all counts, not filtered. The dimensions of the counts table are listed below. Row=genes, Columns=samples
# normal full counts
x <- data.matrix(cpm(genes))
dim(x)
x <- x+1
log_x<-log(x)
names<-colnames(log_x)
pca = prcomp(t(log_x))
summary(pca)
fac = factor(physiology)
colours = function(vec){
  cols=rainbow(length(unique(vec)))
  return(cols[as.numeric(as.factor(vec))])}
#mar.default <- c(5,4,4,2) + 0.1
#par(mar = mar.default + c(0, 4, 0, 0)) 
plot(pca$x[,1:2], 
     col=colours(clade), 
     pch = c(16, 2, 9)[as.numeric(as.factor(physiology))],
     cex=2,
     xlab="PC1",
     ylab="PC2",
     cex.lab=2,
     cex.axis = 2)
#legend(140,100,legend=c("Clade 1","Clade 2","Clade 3"),col=rainbow(length(unique(clade))),cex=0.75, pch=19)
#legend(140,-67,legend=c("Freshwater","Marine"),cex=0.75,pch=c(16, 2, 9))
legend(-75,50,legend=c("Clade 1","Clade 2","Clade 3"),col=rainbow(length(unique(clade))),cex=0.75, pch=19)
legend(-75,25,legend=c("Freshwater","Marine"),cex=0.75,pch=c(16, 2, 9))



meta <- read.table("C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/DE/counts_threshold_new/ExpDesign.txt",stringsAsFactors = FALSE, header = TRUE)
rownames(meta) <- meta$sample
meta <- meta[,-5]
meta$physiology <- gsub('M', 'euryhaline', meta$physiology)
meta$physiology <- gsub('FW', 'stenohaline', meta$physiology)
meta$clade <- gsub('Clade1', 'clade 1', meta$clade)
meta$clade <- gsub('Clade2', 'clade 2', meta$clade)
meta$clade <- gsub('Clade3', 'clade 3', meta$clade)

head(meta)   

pca1 <- pca$x[,1:2]
pca1 <- merge(pca1, meta, by = "row.names")

#png("C:/Users/jmcgirr/Documents/Whitehead_Lab/fun/figs/pca_all_counts.png", height = 5, width = 6.5, units = 'in', res = 600)

ggplot(data = pca1)+
  geom_point(data = pca1, aes(x = PC1, y = PC2, color = physiology, shape = clade), size = 4.5)+
  scale_color_manual(values=c(blu,red))+
  scale_shape_manual(values=c(1,0,2))+
  xlab("10.37%\n")+
  ylab("8.63%\n")+
  theme_bw()#+
  #ggtitle("all populations and sampling years")

#dev.off()


```

# DE Results

```{r filt, results="show",warning=FALSE}
# -------------------
# coefficient names
# -------------------
colnames(coef(fitRan))
# > colnames(coef(fitRan))
#  [1] "(Intercept)"                               "physiologyM"                               "conditiontransfer"                        
#  [4] "cladeClade2"                               "cladeClade3"                               "physiologyM:conditiontransfer"            
#  [7] "physiologyM:cladeClade2"                   "physiologyM:cladeClade3"                   "conditiontransfer:cladeClade2"            
# [10] "conditiontransfer:cladeClade3"             "physiologyM:conditiontransfer:cladeClade2" "physiologyM:conditiontransfer:cladeClade3"

# ---------------------
# transfer (condition) main effect
# ---------------------
contrasts_salinity_main <- makeContrasts("conditiontransfer", "physiologyM.conditiontransfer","conditiontransfer.cladeClade2","conditiontransfer.cladeClade3","physiologyM.conditiontransfer.cladeClade2","physiologyM.conditiontransfer.cladeClade3",levels=make.names(colnames(coef(fitRan))))
rownames(contrasts_salinity_main) <- colnames(coef(fitRan))
vfit <- contrasts.fit(fitRan, contrasts = contrasts_salinity_main)
vfit <- eBayes(vfit)
main_salinity<-topTableF(vfit,n = Inf, sort.by = "F")
sig_main_salinity <- main_salinity[main_salinity$adj.P.Val<0.05,]
dim(sig_main_salinity)
dim(main_salinity)
# fix rownames to chop .1 off
rownames(main_salinity) <- unlist(strsplit(rownames(main_salinity),split="[.]1"))
ann_salinity <- merge(main_salinity,ann,all=TRUE,by.x = "row.names", by.y  = "ensembl_peptide_id")
ann_salinity <- ann_salinity[!is.na(ann_salinity$adj.P.Val),]
dim(ann_salinity)
ann_salinity <- ann_salinity[order(ann_salinity$adj.P.Val,decreasing = FALSE), ]
rownames(ann_salinity) <- ann_salinity$Row.names
ann_salinity <- ann_salinity[,-1]
ann_salinity$description <- NULL
#write.csv(ann_salinity, file = file.path("C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/DE/counts_threshold_new/DE_limma_FW_v_transfer/DE_results/main_transfer.csv"), quote = F, row.names = T)

# ---------------------
# physiology main effect
# ---------------------
contrasts_physiology_main <- makeContrasts("physiologyM", "physiologyM.cladeClade2","physiologyM.cladeClade3","physiologyM.conditiontransfer","physiologyM.conditiontransfer.cladeClade2","physiologyM.conditiontransfer.cladeClade3",levels=make.names(colnames(coef(fitRan))))
rownames(contrasts_physiology_main) <- colnames(coef(fitRan))
vfit <- contrasts.fit(fitRan, contrasts = contrasts_physiology_main)
vfit <- eBayes(vfit)
main_physiology<-topTableF(vfit,n = Inf, sort.by = "F")
sig_main_physiology <- main_physiology[main_physiology$adj.P.Val<0.05,]
dim(sig_main_physiology)
dim(main_physiology)
# fix rownames to chop .1 off
rownames(main_physiology) <- unlist(strsplit(rownames(main_physiology),split="[.]1"))
ann_physiology <- merge(main_physiology,ann,all=TRUE,by.x = "row.names", by.y  = "ensembl_peptide_id")
ann_physiology <- ann_physiology[!is.na(ann_physiology$adj.P.Val),]
dim(ann_physiology)
ann_physiology <- ann_physiology[order(ann_physiology$adj.P.Val,decreasing = FALSE), ]
rownames(ann_physiology) <- ann_physiology$Row.names
ann_physiology <- ann_physiology[,-1]
ann_physiology$description <- NULL
#write.csv(ann_physiology, file = file.path("C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/DE/counts_threshold_new/DE_limma_FW_v_transfer/DE_results/main_physiology.csv"), quote = F, row.names = T)


# ---------------------
# clade main effect
# ---------------------
contrasts_clade_main <- makeContrasts("cladeClade2","cladeClade3","physiologyM.cladeClade2","physiologyM.cladeClade3","conditiontransfer.cladeClade2","conditiontransfer.cladeClade3","physiologyM.conditiontransfer.cladeClade2","physiologyM.conditiontransfer.cladeClade3",levels=make.names(colnames(coef(fitRan))))
rownames(contrasts_clade_main) <- colnames(coef(fitRan))
vfit <- contrasts.fit(fitRan, contrasts = contrasts_clade_main)
vfit <- eBayes(vfit)
main_clade<-topTableF(vfit,n = Inf, sort.by = "F")
sig_main_clade <- main_clade[main_clade$adj.P.Val<0.05,]
dim(sig_main_clade)
dim(main_clade)
# fix rownames to chop .1 off
rownames(main_clade) <- unlist(strsplit(rownames(main_clade),split="[.]1"))
ann_clade <- merge(main_clade,ann,all=TRUE,by.x = "row.names", by.y  = "ensembl_peptide_id")
ann_clade <- ann_clade[!is.na(ann_clade$adj.P.Val),]
dim(ann_clade)
ann_clade <- ann_clade[order(ann_clade$adj.P.Val,decreasing = FALSE), ]
rownames(ann_clade) <- ann_clade$Row.names
ann_clade <- ann_clade[,-1]
ann_clade$description <- NULL
#write.csv(ann_clade, file = file.path("C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/DE/counts_threshold_new/DE_limma_FW_v_transfer/DE_results/main_clade.csv"), quote = F, row.names = T)


# ---------------------
# three-way interaction
# ---------------------
contrasts_threeway <- makeContrasts("physiologyM.conditiontransfer.cladeClade2", "physiologyM.conditiontransfer.cladeClade3", levels=make.names(colnames(coef(fitRan))))
rownames(contrasts_threeway) <- colnames(coef(fitRan))
vfit <- contrasts.fit(fitRan, contrasts = contrasts_threeway)
vfit <- eBayes(vfit)
main_threeway<-topTableF(vfit,n = Inf, sort.by = "F")
sig_threeway <- main_threeway[main_threeway$adj.P.Val<0.05,]
dim(sig_threeway)
dim(main_threeway)
# fix rownames to chop .1 off
rownames(main_threeway) <- unlist(strsplit(rownames(main_threeway),split="[.]1"))
ann_threeway <- merge(main_threeway,ann,all=TRUE,by.x = "row.names", by.y  = "ensembl_peptide_id")
ann_threeway <- ann_threeway[!is.na(ann_threeway$adj.P.Val),]
dim(ann_threeway)
ann_threeway <- ann_threeway[order(ann_threeway$adj.P.Val,decreasing = FALSE), ]
rownames(ann_threeway) <- ann_threeway$Row.names
ann_threeway <- ann_threeway[,-1]
ann_threeway$description <- NULL
#write.csv(ann_threeway, file = file.path("C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/DE/counts_threshold_new/DE_limma_FW_v_transfer/DE_results/threeway.csv"), quote = F, row.names = T)

# ---------------------
# salinity x clade two-way interaction
# ---------------------
contrasts_salinity_clade_interaction <- makeContrasts("conditiontransfer.cladeClade2","conditiontransfer.cladeClade3","physiologyM.conditiontransfer.cladeClade2","physiologyM.conditiontransfer.cladeClade3",levels=make.names(colnames(coef(fitRan))))
rownames(contrasts_salinity_clade_interaction) <- colnames(coef(fitRan))
vfit <- contrasts.fit(fitRan, contrasts = contrasts_salinity_clade_interaction)
vfit <- eBayes(vfit)
main_salinity_clade_interaction<-topTableF(vfit,n = Inf, sort.by = "F")
sig_salinity_clade_interaction <- main_salinity_clade_interaction[main_salinity_clade_interaction$adj.P.Val<0.05,]
dim(sig_salinity_clade_interaction)
dim(main_salinity_clade_interaction)
# fix rownames to chop .1 off
rownames(main_salinity_clade_interaction) <- unlist(strsplit(rownames(main_salinity_clade_interaction),split="[.]1"))
ann_salinity_clade_interaction <- merge(main_salinity_clade_interaction,ann,all=TRUE,by.x = "row.names", by.y  = "ensembl_peptide_id")
ann_salinity_clade_interaction <- ann_salinity_clade_interaction[!is.na(ann_salinity_clade_interaction$adj.P.Val),]
dim(ann_salinity_clade_interaction)
ann_salinity_clade_interaction <- ann_salinity_clade_interaction[order(ann_salinity_clade_interaction$adj.P.Val,decreasing = FALSE), ]
rownames(ann_salinity_clade_interaction) <- ann_salinity_clade_interaction$Row.names
ann_salinity_clade_interaction <- ann_salinity_clade_interaction[,-1]
ann_salinity_clade_interaction$description <- NULL
#write.csv(ann_salinity_clade_interaction, file = file.path("C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/DE/counts_threshold_new/DE_limma_FW_v_transfer/DE_results/transfer_clade_interaction.csv"), quote = F, row.names = T)

# ---------------------
# salinity x physiology two-way interaction
# ---------------------
contrasts_salinity_physiology_interaction <- makeContrasts("physiologyM.conditiontransfer","physiologyM.conditiontransfer.cladeClade2","physiologyM.conditiontransfer.cladeClade3",levels=make.names(colnames(coef(fitRan))))
rownames(contrasts_salinity_physiology_interaction) <- colnames(coef(fitRan))
vfit <- contrasts.fit(fitRan, contrasts = contrasts_salinity_physiology_interaction)
vfit <- eBayes(vfit)
main_salinity_physiology_interaction<-topTableF(vfit,n = Inf, sort.by = "F")
sig_salinity_physiology_interaction <- main_salinity_physiology_interaction[main_salinity_physiology_interaction$adj.P.Val<0.05,]
dim(sig_salinity_physiology_interaction)
dim(main_salinity_physiology_interaction)
# fix rownames to chop .1 off
rownames(main_salinity_physiology_interaction) <- unlist(strsplit(rownames(main_salinity_physiology_interaction),split="[.]1"))
ann_salinity_physiology_interaction <- merge(main_salinity_physiology_interaction,ann,all=TRUE,by.x = "row.names", by.y  = "ensembl_peptide_id")
ann_salinity_physiology_interaction <- ann_salinity_physiology_interaction[!is.na(ann_salinity_physiology_interaction$adj.P.Val),]
dim(ann_salinity_physiology_interaction)
ann_salinity_physiology_interaction <- ann_salinity_physiology_interaction[order(ann_salinity_physiology_interaction$adj.P.Val,decreasing = FALSE), ]
rownames(ann_salinity_physiology_interaction) <- ann_salinity_physiology_interaction$Row.names
ann_salinity_physiology_interaction <- ann_salinity_physiology_interaction[,-1]
ann_salinity_physiology_interaction$description <- NULL
#write.csv(ann_salinity_physiology_interaction, file = file.path("C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/DE/counts_threshold_new/DE_limma_FW_v_transfer/DE_results/transfer_physiology_interaction.csv"), quote = F, row.names = T)

# ---------------------
# physiology x clade two-way interaction
# ---------------------
contrasts_clade_physiology_interaction <- makeContrasts("physiologyM.cladeClade2","physiologyM.cladeClade3","physiologyM.conditiontransfer.cladeClade2","physiologyM.conditiontransfer.cladeClade3",levels=make.names(colnames(coef(fitRan))))
rownames(contrasts_clade_physiology_interaction) <- colnames(coef(fitRan))
vfit <- contrasts.fit(fitRan, contrasts = contrasts_clade_physiology_interaction)
vfit <- eBayes(vfit)
main_clade_physiology_interaction<-topTableF(vfit,n = Inf, sort.by = "F")
sig_clade_physiology_interaction <- main_clade_physiology_interaction[main_clade_physiology_interaction$adj.P.Val<0.05,]
dim(sig_clade_physiology_interaction)
dim(main_clade_physiology_interaction)
# fix rownames to chop .1 off
rownames(main_clade_physiology_interaction) <- unlist(strsplit(rownames(main_clade_physiology_interaction),split="[.]1"))
ann_clade_physiology_interaction <- merge(main_clade_physiology_interaction,ann,all=TRUE,by.x = "row.names", by.y  = "ensembl_peptide_id")
ann_clade_physiology_interaction <- ann_clade_physiology_interaction[!is.na(ann_clade_physiology_interaction$adj.P.Val),]
dim(ann_clade_physiology_interaction)
ann_clade_physiology_interaction <- ann_clade_physiology_interaction[order(ann_clade_physiology_interaction$adj.P.Val,decreasing = FALSE), ]
rownames(ann_clade_physiology_interaction) <- ann_clade_physiology_interaction$Row.names
ann_clade_physiology_interaction <- ann_clade_physiology_interaction[,-1]
ann_clade_physiology_interaction$description <- NULL
#write.csv(ann_clade_physiology_interaction, file = file.path("C:/Users/jmcgirr/Documents/GitHub/RNAseq_17killifish_osmotic/DE/counts_threshold_new/DE_limma_FW_v_transfer/DE_results/clade_physiology_interaction.csv"), quote = F, row.names = T)



#======================

#-------------------------------------
# significant genes
#-------------------------------------

dim(sig_clade_physiology_interaction)
dim(sig_salinity_physiology_interaction)
dim(sig_salinity_clade_interaction)
dim(sig_threeway)
dim(sig_main_clade)
dim(sig_main_physiology)
dim(sig_main_salinity)

#> dim(sig_clade_physiology_interaction)
#[1] 1315    8
#> dim(sig_salinity_physiology_interaction)
#[1] 739   7
#> dim(sig_salinity_clade_interaction)
#[1] 224   8
#> dim(sig_threeway)
#[1] 21  6
#> dim(sig_main_clade)
#[1] 4115   12
#> dim(sig_main_physiology)
#[1] 3357   10
#> dim(sig_main_salinity)
#[1] 3228   10

```



## summary table
```{r filt, results="show",warning=FALSE}


# make a summary table for Andrew:
# ---------------------------------
# cols:
# Gene ID
# gene name
# gene descriptions
# adj.P.Val.Clade
# adj.P.Val.Salinity
# adj.P.Val.Physiology
# adj.P.Val.Salinity.Physiology
# adj.P.Val.Physiology.Clade
# adj.P.Val.Salinity.Clade
# adj.P.Val.3way
# norm counts
# ---------------------------------

new <- merge(ann_clade,ann_physiology,by= "row.names")
colnames(new)[13] <- "adj.P.Val.Clade"
colnames(new)[28] <- "adj.P.Val.Physiology"
new <- new[ -c(2:12, 19:27,29:33) ]
colnames(new)[1] <- "ID"
new <- merge(new,ann_salinity,by.x= "ID",by.y="row.names")
colnames(new)[18] <- "adj.P.Val.Salinity"
new <- new[ -c(9:17,19:23) ]
new <- new[,c(1,3:7,2,8:9)]
new <- merge(new,ann_clade_physiology_interaction,by.x= "ID",by.y="row.names")
colnames(new)[17] <- "adj.P.Val.Physiology.Clade"
new <- new[ -c(10:16,18:22) ]
new <- merge(new,ann_salinity_physiology_interaction,by.x= "ID",by.y="row.names")
colnames(new)[17] <- "adj.P.Val.Salinity.Physiology"
new <- new[ -c(11:16,18:22) ]
new <- merge(new,ann_salinity_clade_interaction,by.x= "ID",by.y="row.names")
colnames(new)[19] <- "adj.P.Val.Salinity.Clade"
new <- new[ -c(12:18,20:24) ]
new <- merge(new,ann_threeway,by.x= "ID",by.y="row.names")
colnames(new)[18] <- "adj.P.Val.3way"
new <- new[ -c(13:17,19:23) ]
colnames(new)[2] <- "ensembl_transcript_id"
colnames(new)[3] <- "ensembl_gene_id"
colnames(new)[4] <- "gene_biotype"
colnames(new)[5] <- "external_gene_name"
colnames(new)[6] <- "description"
new <- merge(new,norm_counts,by.x= "ID",by.y="row.names")
# ---------------------------------
# mean expression each species BW
# mean expression each species FW
# x 14
new$F.rath.BW.mean <- log2(rowMeans(new[c(14:16)], na.rm=TRUE)+1)
new$F.rath.FW.mean <- log2(rowMeans(new[c(17:19)], na.rm=TRUE)+1)
new$F.grandis.BW.mean <- log2(rowMeans(new[c(20:22)], na.rm=TRUE)+1)
new$F.grandis.FW.mean <- log2(rowMeans(new[c(23:25)], na.rm=TRUE)+1)
new$F.notatus.BW.mean <- log2(rowMeans(new[c(26:28)], na.rm=TRUE)+1)
new$F.notatus.FW.mean <- log2(rowMeans(new[c(29:31)], na.rm=TRUE)+1)
new$F.parv.BW.mean <- log2(rowMeans(new[c(32:34)], na.rm=TRUE)+1)
new$F.parv.FW.mean <- log2(rowMeans(new[c(35:37)], na.rm=TRUE)+1)
new$L.good.BW.mean <- log2(rowMeans(new[c(38:40)], na.rm=TRUE)+1)
new$L.good.FW.mean <- log2(rowMeans(new[c(41:43)], na.rm=TRUE)+1)
new$F.oli.BW.mean <- log2(rowMeans(new[c(44:46)], na.rm=TRUE)+1)
new$F.oli.FW.mean <- log2(rowMeans(new[c(47:49)], na.rm=TRUE)+1)
new$L.parv.BW.mean <- log2(rowMeans(new[c(50:52)], na.rm=TRUE)+1)
new$L.parv.FW.mean <- log2(rowMeans(new[c(53:55)], na.rm=TRUE)+1)
new$F.hetPP.BW.mean <- log2(rowMeans(new[c(56:58)], na.rm=TRUE)+1)
new$F.hetPP.FW.mean <- log2(rowMeans(new[c(59:61)], na.rm=TRUE)+1)
new$F.sim.BW.mean <- log2(rowMeans(new[c(62:64)], na.rm=TRUE)+1)
new$F.sim.FW.mean <- log2(rowMeans(new[c(65:67)], na.rm=TRUE)+1)
new$F.dia.BW.mean <- log2(rowMeans(new[c(68:69)], na.rm=TRUE)+1)
new$F.dia.FW.mean <- log2(rowMeans(new[c(70:71)], na.rm=TRUE)+1)
new$F.chry.BW.mean <- log2(rowMeans(new[c(72:74)], na.rm=TRUE)+1)
new$F.chry.FW.mean <- log2(rowMeans(new[c(72:77)], na.rm=TRUE)+1)
new$A.xen.BW.mean <- log2(rowMeans(new[c(78:80)], na.rm=TRUE)+1)
new$A.xen.FW.mean <- log2(rowMeans(new[c(81:83)], na.rm=TRUE)+1)
new$F.cat.BW.mean <- log2(rowMeans(new[c(84:86)], na.rm=TRUE)+1)
new$F.cat.FW.mean <- log2(rowMeans(new[c(87:88)], na.rm=TRUE)+1)
new$F.hetPL.BW.mean <- log2(rowMeans(new[c(89:91)], na.rm=TRUE)+1)
new$F.hetPL.FW.mean <- log2(rowMeans(new[c(92:94)], na.rm=TRUE)+1)
# ---------------------------------
# log2FC FW=0 - each species - zero normalized for salinity
# log2FC BW=15 - each species

# subtract FW from itself = 0
# subtract FW from BW = 15

new$F.rath.0 <- (new$F.rath.FW.mean - new$F.rath.FW.mean) 
new$F.rath.15 <- (new$F.rath.BW.mean - new$F.rath.FW.mean)
new$F.grandis.0 <- (new$F.grandis.FW.mean - new$F.grandis.FW.mean)
new$F.grandis.15 <- (new$F.grandis.BW.mean - new$F.grandis.FW.mean)
new$F.notatus.0 <- (new$F.notatus.FW.mean - new$F.notatus.FW.mean)
new$F.notatus.15 <- (new$F.notatus.BW.mean - new$F.notatus.FW.mean)
new$F.parv.0 <- (new$F.parv.FW.mean - new$F.parv.FW.mean)
new$F.parv.15 <- (new$F.parv.BW.mean - new$F.parv.FW.mean)
new$L.good.0 <- (new$L.good.FW.mean - new$L.good.FW.mean)
new$L.good.15 <- (new$L.good.BW.mean - new$L.good.FW.mean)
new$F.oli.0 <- (new$F.oli.FW.mean - new$F.oli.FW.mean)  
new$F.oli.15 <- (new$F.oli.BW.mean - new$F.oli.FW.mean)  
new$L.parv.0 <- (new$L.parv.FW.mean - new$L.parv.FW.mean) 
new$L.parv.15 <- (new$L.parv.BW.mean - new$L.parv.FW.mean)
new$F.hetPP.0 <- (new$F.hetPP.FW.mean - new$F.hetPP.FW.mean) 
new$F.hetPP.15 <- (new$F.hetPP.BW.mean - new$F.hetPP.FW.mean)
new$F.sim.0 <- (new$F.sim.FW.mean - new$F.sim.FW.mean)
new$F.sim.15 <- (new$F.sim.BW.mean - new$F.sim.FW.mean)
new$F.dia.0 <- (new$F.dia.FW.mean - new$F.dia.FW.mean)
new$F.dia.15 <- (new$F.dia.BW.mean - new$F.dia.FW.mean)
new$F.chry.0 <- (new$F.chry.FW.mean - new$F.chry.FW.mean)
new$F.chry.15 <- (new$F.chry.BW.mean - new$F.chry.FW.mean)
new$A.xen.0 <- (new$A.xen.FW.mean - new$A.xen.FW.mean)
new$A.xen.15 <- (new$A.xen.BW.mean - new$A.xen.FW.mean)
new$F.cat.0 <- (new$F.cat.FW.mean - new$F.cat.FW.mean)
new$F.cat.15 <- (new$F.cat.BW.mean - new$F.cat.FW.mean)
new$F.hetPL.0 <- (new$F.hetPL.FW.mean - new$F.hetPL.FW.mean)
new$F.hetPL.15 <- (new$F.hetPL.BW.mean - new$F.hetPL.FW.mean)
# ---------------------------------
# clade-specific grand mean normalization
new$Clade1mean <- log2(rowMeans(new[c(14:25,56:61,84:94,68:71)], na.rm=TRUE)+1)
new$Clade2mean <- log2(rowMeans(new[c(32:43,50:55)], na.rm=TRUE)+1)
new$Clade3mean <- log2(rowMeans(new[c(78:83,72:77,62:67,26:31,44:49)], na.rm=TRUE)+1)

# subtract each species/condition for mean
# clade 1
new$F.rath.0.clade <- (new$F.rath.FW.mean - new$Clade1mean) 
new$F.rath.15.clade <- (new$F.rath.BW.mean - new$Clade1mean)
new$F.grandis.0.clade <- (new$F.grandis.FW.mean - new$Clade1mean)
new$F.grandis.15.clade <- (new$F.grandis.BW.mean - new$Clade1mean)
new$F.hetPP.0.clade <- (new$F.hetPP.FW.mean - new$Clade1mean) 
new$F.hetPP.15.clade <- (new$F.hetPP.BW.mean - new$Clade1mean)
new$F.dia.0.clade <- (new$F.dia.FW.mean - new$Clade1mean)
new$F.dia.15.clade <- (new$F.dia.BW.mean - new$Clade1mean)
new$F.hetPL.0.clade <- (new$F.hetPL.FW.mean - new$Clade1mean)
new$F.hetPL.15.clade <- (new$F.hetPL.BW.mean - new$Clade1mean)
new$F.cat.0.clade <- (new$F.cat.FW.mean - new$Clade1mean)
new$F.cat.15.clade <- (new$F.cat.BW.mean - new$Clade1mean)
#clade 2
new$F.parv.0.clade <- (new$F.parv.FW.mean - new$Clade2mean)
new$F.parv.15.clade <- (new$F.parv.BW.mean - new$Clade2mean)
new$L.good.0.clade <- (new$L.good.FW.mean - new$Clade2mean)
new$L.good.15.clade <- (new$L.good.BW.mean - new$Clade2mean)
new$L.parv.0.clade <- (new$L.parv.FW.mean - new$Clade2mean) 
new$L.parv.15.clade <- (new$L.parv.BW.mean - new$Clade2mean)
# clade 3
new$F.oli.0.clade <- (new$F.oli.FW.mean - new$Clade3mean)  
new$F.oli.15.clade <- (new$F.oli.BW.mean - new$Clade3mean)  
new$F.sim.0.clade <- (new$F.sim.FW.mean - new$Clade3mean)
new$F.sim.15.clade <- (new$F.sim.BW.mean - new$Clade3mean)
new$F.chry.0.clade <- (new$F.chry.FW.mean - new$Clade3mean)
new$F.chry.15.clade <- (new$F.chry.BW.mean - new$Clade3mean)
new$A.xen.0.clade <- (new$A.xen.FW.mean - new$Clade3mean)
new$A.xen.15.clade <- (new$A.xen.BW.mean - new$Clade3mean)
new$F.notatus.0.clade <- (new$F.notatus.FW.mean - new$Clade3mean)
new$F.notatus.15.clade <- (new$F.notatus.BW.mean - new$Clade3mean)





#write.csv(new,file = file.path("~/Documents/UCDavis/Whitehead/kfish_expression_July2019/kfishosmotictolerance_Summary.Table.csv"), quote = F, row.names = T)

#write.table(new,"~/Documents/UCDavis/Whitehead/kfish_expression_July2019/kfishosmotictolerance_Summary.Table.tsv",sep="\t",quote=F,row.names=FALSE)

dim(new)

#-------------------
# Set sample order
#-------------------

sample_phys <-c("F.grandis.0.clade","F.grandis.15.clade","F.hetPL.0.clade","F.hetPL.15.clade","F.hetPP.0.clade","F.hetPP.15.clade","F.dia.0.clade","F.dia.15.clade","F.cat.0.clade","F.cat.15.clade","F.rath.0.clade","F.rath.15.clade","F.parv.0.clade","F.parv.15.clade","L.parv.0.clade","L.parv.15.clade","L.good.0.clade","L.good.15.clade","A.xen.0.clade","A.xen.15.clade","F.chry.0.clade","F.chry.15.clade","F.sim.0.clade","F.sim.15.clade","F.notatus.0.clade","F.notatus.15.clade","F.oli.0.clade","F.oli.15.clade")

sample <- c("F.grandis.0","F.grandis.15","F.hetPL.0","F.hetPL.15","F.hetPP.0","F.hetPP.15","F.dia.0","F.dia.15","F.cat.0","F.cat.15","F.rath.0","F.rath.15","F.parv.0","F.parv.15","L.parv.0","L.parv.15","L.good.0","L.good.15","A.xen.0","A.xen.15","F.chry.0","F.chry.15","F.sim.0","F.sim.15","F.notatus.0","F.notatus.15","F.oli.0","F.oli.15")

clade <- c("clade1","clade1","clade1","clade1","clade1","clade1","clade1","clade1","clade1","clade1","clade1","clade1","clade2","clade2","clade2","clade2","clade2","clade2","clade3","clade3","clade3","clade3","clade3","clade3","clade3","clade3","clade3","clade3")
physiology <- c("M","M","M","M","M","M","M","M","F","F","F","F","M","M","M","M","F","F","M","M","M","M","M","M","F","F","F","F")
salinity <- c("0ppt","15ppt","0ppt","15ppt","0ppt","15ppt","0ppt","15ppt","0ppt","15ppt","0ppt","15ppt","0ppt","15ppt","0ppt","15ppt","0ppt","15ppt","0ppt","15ppt","0ppt","15ppt","0ppt","15ppt","0ppt","15ppt","0ppt","15ppt")
sample_label_df <- data.frame(salinity,physiology,clade)
rownames(sample_label_df) <- colnames(h)

```


# Heatmaps

```{r filt, results="show",warning=FALSE}
# adj.P.Val.Clade
# adj.P.Val.Salinity
# adj.P.Val.Physiology
# adj.P.Val.Salinity.Physiology
# adj.P.Val.Physiology.Clade
# adj.P.Val.Salinity.Clade
# adj.P.Val.3way

# conserved salinity
cons_salinity <- filter(new, adj.P.Val.Salinity<=0.05 & adj.P.Val.Salinity.Physiology>0.1 & adj.P.Val.3way>0.1)
h <- cons_salinity[,c(123:150)]
rownames(h)<-cons_salinity$ID
h <- h[,sample]
rownames(sample_label_df) <- colnames(h)

# main effect of physiology, with no physiology*condition interaction
# conserved physiology
cons_physiology <- filter(new, adj.P.Val.Physiology<=0.05 & adj.P.Val.Salinity.Physiology>0.1 & adj.P.Val.3way>0.1)
h <- cons_physiology[,c(154:181)]
rownames(h)<-cons_physiology$ID
h <- h[,sample_phys]
rownames(sample_label_df) <- colnames(h)

# Divergent clade effect 
cons_clade <- filter(new, adj.P.Val.Clade<=0.05 & adj.P.Val.Salinity<=0.05)
h <- cons_clade[,c(123:150)]
rownames(h)<-cons_clade$ID
h <- h[,sample]
rownames(sample_label_df) <- colnames(h)

# salinity-physiology interaction
phys_salinity <- filter(new, adj.P.Val.Salinity.Physiology<=0.05 & adj.P.Val.3way>0.1)
h <- phys_salinity[,c(123:150)]
rownames(h)<-phys_salinity$ID
h <- h[,sample]
rownames(sample_label_df) <- colnames(h)

# salinity-clade interaction
salinity_clade <- filter(new, adj.P.Val.Salinity.Clade<=0.05 & adj.P.Val.3way>0.1)
h <- salinity_clade[,c(123:150)]
rownames(h)<-salinity_clade$ID
h <- h[,sample]
rownames(sample_label_df) <- colnames(h)

# physiolgy-clade interaction
physiology_clade <- filter(new, adj.P.Val.Physiology.Clade<=0.05 & adj.P.Val.3way>0.1)
h <- physiology_clade[,c(123:150)]
rownames(h)<-physiology_clade$ID
h <- h[,sample]
rownames(sample_label_df) <- colnames(h)

# threeway interaction
threeway <- filter(new, adj.P.Val.3way<=0.05)
h <- threeway[,c(123:150)]
rownames(h)<-threeway$ID
h <- h[,sample]
rownames(sample_label_df) <- colnames(h)



#--------------------------
# Set heatmap color scale
#--------------------------

my.breaks <- c(seq(-2, 0, by=0.1), seq(0.1, 2, by=0.1))
my.colors <- c(colorRampPalette(colors = c("#00FFFF", "black"))(length(my.breaks)/2), colorRampPalette(colors = c("black", "yellow"))(length(my.breaks)/2))

annotation_colors = list(salinity = c("0ppt"="red", "15ppt"="blue"),physiology = c("M"="blue", "F"="red"),clade = c("clade1"="green", "clade2"="orange", "clade3"="purple"))

# ------------------
# Run heatmap
# ------------------

out <- pheatmap(h, cluster_rows = TRUE,
         clustering_distance_rows = "correlation",
         cluster_cols = F,
         annotation_col = sample_label_df,
         show_rownames=F,
         cutree_rows = 3,
         color = my.colors,
         annotation_colors = annotation_colors,
         breaks = my.breaks,
         gaps_col = c(12,18)
)

# subset
cut2groups <- data.frame(sort(cutree(out$tree_row, k=2)))
colnames(cut2groups)[1] <- "group"

phys_salinity$ID <- as.character(phys_salinity$ID)
salinity_phys <- merge(phys_salinity,cut2groups,by.x="ID",by.y="row.names")
group1up <- salinity_phys[salinity_phys$group == 1,]
group2down <- salinity_phys[salinity_phys$group == 2,]

cons_salinity_groups <- merge(cons_salinity,cut2groups,by.x="ID",by.y="row.names")
group1up <- cons_salinity_groups[cons_salinity_groups$group == 1,]
group2down <- cons_salinity_groups[cons_salinity_groups$group == 2,]

h <- group1up[,c(123:150)]
rownames(h)<-group1up$ID
h <- h[,sample]
rownames(sample_label_df) <- colnames(h)
out <- pheatmap(h, cluster_rows = TRUE,
                clustering_distance_rows = "correlation",
                cluster_cols = F,
                annotation_col = sample_label_df,
                show_rownames=F,
                color = my.colors,
                annotation_colors = annotation_colors,
                breaks = my.breaks,
                gaps_col = c(12,18)
)

# visually group looks up (even though not all are)
up <- cons_salinity[,c(123:150)]
rownames(up)<-cons_salinity$ID
# only 15
up <- up[,c(2,4,6,8,10,12,14,16,18,20,22,24,26,28)]
up <- up[apply(up > 0, 1, all), ]
# only 14
upID <- rownames(up)
new_up <- new[new$ID %in% upID,]
new_up <- merge(new_up,cut2groups,by.x="ID",by.y="row.names")

down <- cons_salinity[,c(123:150)]
rownames(down)<-cons_salinity$ID
down <- down[,c(2,4,6,8,10,12,14,16,18,20,22,24,26,28)]
down <- down[apply(down < 0, 1, all), ]
# only 4
downID <- rownames(down)
new_down <- new[new$ID %in% downID,]
new_down <- merge(new_down,cut2groups,by.x="ID",by.y="row.names")


circ_genes
# these go for GO analysis
length(group1up$ID)
# 272
length(group2down$ID)
# 310
dim(cons_salinity)
# 582


# cons_physiology

cut2groups <- data.frame(sort(cutree(out$tree_row, k=3)))
colnames(cut2groups)[1] <- "group"

cons_physiology$ID <- as.character(cons_physiology$ID)
physiology_cons <- merge(cons_physiology,cut2groups,by.x="ID",by.y="row.names")
group1up <- physiology_cons[physiology_cons$group== 1,]
group2down <- physiology_cons[physiology_cons$group == 2,]
group3 <- physiology_cons[physiology_cons$group == 3,]
dim(group1up)
dim(group2down)
dim(group3)



# sort by adj.P.Val.Physiology
# get top 20
# make table of genes and GO terms
cons_physiology_sort <- group3[order(group3$adj.P.Val.Physiology),]
cons_physiology_50 <- cons_physiology_sort[1:50,]
  
# sort by phys-sal
# get top 20
  
# sort by adj.P.Val.Salinity
# get top 20
# make table of genes and GO terms
cons_salinity_sort <- cons_salinity
```
